---
title: "Cause of Death anaylsis"
author:
- 'Projektpartner: Prof. Dr. Li'
- 'Betreuer: Helen Alber'
date: "2025-03-19"
output:
  slidy_presentation: default
  ioslides_presentation: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Packages, include=FALSE}
library(here)
library(ggplot2)
library(readr)
library(dplyr)
library(patchwork)
options(scipen=999) #normale Zahlen statt 2e+05 oder so
```

```{r Univariate Analysis, include=FALSE}
plot1 <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(title = "Sepal Length vs Sepal Width in Iris Dataset") +
  theme_minimal()
```

```{r Bivariate Analysis, include=FALSE}

```

```{r Hypothesis Tests, include=FALSE}

```

```{r Model, include=FALSE}

```


## Research Question
- Exploratory Role
- main question is the influence of the socioeconomic status on the mortality rate of certain diseases
- Goal: build, if possible, one model that returns the probability of dying from a disease given socioeconomic status and other variables that may have an influence on the mortality.


## About our Data
- Data is provided through data querying system run by the CDC (Centers for Disease Control and Prevention): wonder.cdc.gov
- We use the dataset from 2018-2023 focused on the underlying causes of death
- datasets are provided pre-grouped
- The underlying causes of death are defined and operationalised according to ICD-10 and are to be analysed as mortality rates as a variable of geographical areas
- The socioeconomic status is operationalised by the highest attained education
- Population size is mostly, but not always given
- Varibales of Interest: Year, Year&Month, Age Groups, Sex, Single Race 6, Education, Urbanization level, States


## What Methodology are we using?
- Univariate Analysis
- Bivariate Analysis
- Hypothesis Tests
- Creation of a Model

## Univariate Analysis
- Exploring Variables across Top 5 Cause of Deaths & Total Deaths one at a time
- Variables: Year, Year&Month, Age Groups, Sex, Single Race 6, Education, Urbanization level, States

## Example slide / Chosen Plots...
```{r Plot1, echo=FALSE}
plot1
```

