```{r Packages}
library(dplyr)
library(readxl)
library(tidyr)
library(broom)
library(survival)
library(ggplot2)
```


What we need in the future:
- A Population so that we can model individual Risk instead of the Counts which are heavily influenced by Population size
- Deaths per 100k Population column which we use for our models
- Add "other" to make it the reference category
```{r load Data}
#load Top_5_Death dataset first
df <- Top_5_Death
```

We need individual data points for AFT Models
--> We split up the death count into individual data points, so that a data point represents 100 Deaths, remove the reduction for final model
- Then we modify the Data, so that it is compatible with AFT Models
```{r prepare Data}
reduced_deaths <- df %>%
    mutate(Deaths = Deaths / 100) %>%
    mutate(Deaths = round(Deaths))

expanded_deaths <- reduced_deaths %>%
  rename(
    Age = `Single-Year Ages Code`,
    CoD = `Cause of Death`
  ) %>%
  mutate(
    Urbanization = as.factor(Urbanization),
    CoD = as.factor(CoD),
    Sex = as.factor(Sex)
  ) %>%
  uncount(weights = Deaths) %>%
  mutate(status = 1)

aft_data <- expanded_deaths %>%
  filter(!is.na(Age) & Age > 0)
```

```{r Models, all interactions}
# Weibull
aft_weibull <- survreg(Surv(Age) ~ Sex * CoD + Sex * Urbanization + CoD*Urbanization, data = aft_data, dist = "weibull")

# Log-normal
aft_lognormal <- survreg(Surv(Age) ~ Sex * CoD + Sex * Urbanization + CoD*Urbanization, data = aft_data, dist = "lognormal")

# Log-logistic
aft_loglogistic <- survreg(Surv(Age) ~ Sex * CoD + Sex * Urbanization + CoD*Urbanization, data = aft_data, dist = "loglogistic")

# Exponential
aft_exponential <- survreg(Surv(Age) ~ Sex * CoD + Sex * Urbanization + CoD*Urbanization, data = aft_data, dist = "exponential")

#show the reference Category
levels(aft_data$Sex)[1]
levels(aft_data$CoD)[1]
levels(aft_data$Urbanization)[1]

#lognormal has the correct intercept
```
```{r Best Model}
#compare with AIC
AIC(aft_weibull, aft_lognormal, aft_loglogistic, aft_exponential)

#compare log.likelihoods
logLik(aft_weibull)
logLik(aft_lognormal)
logLik(aft_loglogistic)
logLik(aft_exponential)

#Weibull is best
```

later: do this for every distribution
```{r Other Weibull Models}
# aft_weibull <- survreg(Surv(Age) ~ Sex * CoD + Sex * Urbanization + CoD*Urbanization, data = aft_data, dist = "weibull")
aft_weibull2 <- survreg(Surv(Age) ~ Sex + CoD + Urbanization, data = aft_data, dist = "weibull")
aft_weibull3 <- survreg(Surv(Age) ~ Sex * CoD + CoD*Urbanization, data = aft_data, dist = "weibull")
aft_weibull4 <- survreg(Surv(Age) ~ Sex * Urbanization, data = aft_data, dist = "weibull")

AIC(aft_weibull, aft_weibull2, aft_weibull3, aft_weibull4) #aft_weibull3 might be the best due to lower risk of overfitting and little difference in AIC
```


```{r Coefficients & Numbers}
summary(aft_weibull3)
exp(coef(summary(aft_weibull3)))

#To analyse differences between predictions
predict_diff <- function(sex1, sex2, cod, urb) {
  predict(aft_weibull3, newdata = data.frame(Sex = sex1, CoD = cod, Urbanization = urb)) -
    predict(aft_weibull3, newdata = data.frame(Sex = sex2, CoD = cod, Urbanization = urb))
}
# Example:
predict_diff("Male", "Female", "Malignant neoplasms", "Small Metro")

#Get expected Age:
predicted_age <- function(sex, cod, urb, data = aft_data) {
predict(aft_weibull3, newdata = data.frame(
  Sex = "Female",
  CoD = "Diseases of heart",
  Urbanization = "Large Central Metro"
))
}
#Validate on actual data:
get_observed_age_summary <- function(sex, cod, urb, data = aft_data) {
  data %>%
    filter(
      Sex == sex,
      CoD == cod,
      Urbanization == urb
    ) %>%
    summarise(
      mean_age = mean(Age),
      median_age = median(Age),
      count = n()
    )
}
predicted_age(sex, cod, urb, data = aft_data)
get_observed_age_summary(sex = "Female", cod = "Malignant neoplasms", urb = "Small Metro")
#not that great, test with other distributions and adjusted Population
```



```{r Residual Plots}
#Deviance residuals
plot(residuals(aft_weibull3), main = "Deviance Residuals")
abline(h = 0, col = "red", lty = 2)

#Standardized Lg Residuals:
my_resid <- log(aft_data$Age) - aft_weibull3$linear.predictors
scale_param <- aft_weibull3$scale  # Extracts the scale (σ) from the model
my_std_resid <- my_resid / scale_param
plot(my_std_resid, main = "Standardized Log-Residuals")
abline(h = 0, col = "red", lty = 2)


#QQ Plot of Residuals
qqnorm(my_resid, main = "QQ Plot of Log-Residuals")
qqline(my_resid)
qqnorm(residuals(aft_weibull3), main = "QQ Plot of Deviance Residuals")
qqline(residuals(aft_weibull3))


#Predicted vs Observed
pred <- predict(aft_weibull3, type = "response")  # Predicted Age
obs <- aft_data$Age
plot(pred, obs, xlab = "Predicted Age", ylab = "Observed Age", main = "Predicted vs. Observed")
abline(0, 1, col = "red")

#Residuals by Group
aft_data$resid <- residuals(aft_weibull3)
ggplot(aft_data, aes(x = 1:nrow(aft_data), y = resid, color = Sex)) +
  geom_point(alpha = 0.5, size = 0.3) +
  theme_minimal() +
  labs(title = "Residuals by Sex", x = "Observation", y = "Residual")

#Histogram of Residuals
hist(my_resid, breaks = 100, main = "Histogram of Log-Residuals", xlab = "Residual")

#Deviance Residual Plot, showing grouped residuals
aft_data$resid <- residuals(aft_weibull3)
ggplot(aft_data, aes(x = 1:nrow(aft_data), y = resid, color = CoD)) +
  geom_point(alpha = 0.5, size = 0.3) +
  theme_minimal() +
  labs(title = "Residuals by Cause of Death", x = "Observation", y = "Residual")

#big problems with multiple death causes, makes sense, as different death causes are affected differently by socio-demographic factors
# => Split them up
```
```{r Residuals for loglogistic}
#Deviance residuals
plot(residuals(aft_loglogistic), main = "Deviance Residuals")
abline(h = 0, col = "red", lty = 2)

#Raw Lg Residuals:
my_resid <- log(aft_data$Age) - aft_loglogistic$linear.predictors
plot(my_resid, main = "Raw Log-Residuals")
abline(h = 0, col = "red", lty = 2)

#QQ Plot of Residuals
qqnorm(my_resid, main = "QQ Plot of Log-Residuals")
qqline(my_resid)
qqnorm(residuals(aft_loglogistic), main = "QQ Plot of Deviance Residuals")
qqline(residuals(aft_loglogistic))


#Predicted vs Observed
pred <- predict(aft_loglogistic, type = "response")  # Predicted Age
obs <- aft_data$Age
plot(pred, obs, xlab = "Predicted Age", ylab = "Observed Age", main = "Predicted vs. Observed")
abline(0, 1, col = "red")

#Residuals by Group
aft_data$resid <- residuals(aft_loglogistic)
ggplot(aft_data, aes(x = 1:nrow(aft_data), y = resid, color = Sex)) +
  geom_point(alpha = 0.5, size = 0.3) +
  theme_minimal() +
  labs(title = "Residuals by Sex", x = "Observation", y = "Residual")

#Histogram of Residuals
hist(my_resid, breaks = 100, main = "Histogram of Log-Residuals", xlab = "Residual")

#Deviance Residual Plot, showing grouped residuals
aft_data$resid <- residuals(aft_loglogistic)
ggplot(aft_data, aes(x = 1:nrow(aft_data), y = resid, color = CoD)) +
  geom_point(alpha = 0.5, size = 0.3) +
  theme_minimal() +
  labs(title = "Residuals by Cause of Death", x = "Observation", y = "Residual")
```

```{r Can we fix it by splitting up death groups into multiple models?}
#Accidents
aft_accidents_weibull <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Accidents (unintentional injuries)"),
  dist = "weibull"
)
aft_accidents_exponential <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Accidents (unintentional injuries)"),
  dist = "exponential"
)
aft_accidents_lognormal <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Accidents (unintentional injuries)"),
  dist = "lognormal"
)
aft_accidents_loglogistic <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Accidents (unintentional injuries)"),
  dist = "loglogistic"
)

# Cerebro
aft_cerebro_weibull <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Cerebrovascular diseases"),
  dist = "weibull"
)
aft_cerebro_exponential <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Cerebrovascular diseases"),
  dist = "exponential"
)
aft_cerebro_lognormal <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Cerebrovascular diseases"),
  dist = "lognormal"
)
aft_cerebro_loglogistic <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Cerebrovascular diseases"),
  dist = "loglogistic"
)

#Covid
aft_covid_weibull <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "COVID-19"),
  dist = "weibull"
)
aft_covid_exponential <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "COVID-19"),
  dist = "exponential"
)
aft_covid_lognormal <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "COVID-19"),
  dist = "lognormal"
)
aft_covid_loglogistic <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "COVID-19"),
  dist = "loglogistic"
)

#Heart Disease
aft_heart_weibull <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Diseases of heart"),
  dist = "weibull"
)
aft_heart_exponential <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Diseases of heart"),
  dist = "exponential"
)
aft_heart_lognormal <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Diseases of heart"),
  dist = "lognormal"
)
aft_heart_loglogistic <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Diseases of heart"),
  dist = "loglogistic"
)

#Malignant Neoplasms
aft_neoplasms_weibull <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Malignant neoplasms"),
  dist = "weibull"
)
aft_neoplasms_exponential <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Malignant neoplasms"),
  dist = "exponential"
)
aft_neoplasms_lognormal <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Malignant neoplasms"),
  dist = "lognormal"
)
aft_neoplasms_loglogistic <- survreg(
  Surv(Age) ~ Sex * Urbanization,
  data = aft_data %>% filter(CoD == "Malignant neoplasms"),
  dist = "loglogistic"
)
```

```{r best Models}
AIC(aft_accidents_weibull, aft_accidents_exponential, aft_accidents_lognormal, aft_accidents_loglogistic)
AIC(aft_cerebro_weibull, aft_cerebro_exponential, aft_cerebro_lognormal, aft_cerebro_loglogistic)
AIC(aft_covid_weibull, aft_covid_exponential, aft_covid_lognormal, aft_covid_loglogistic)
AIC(aft_heart_weibull, aft_heart_exponential, aft_heart_lognormal, aft_heart_loglogistic)
AIC(aft_neoplasms_weibull, aft_neoplasms_exponential, aft_neoplasms_lognormal, aft_neoplasms_loglogistic)
#Weibull is best for everyone
```

```{r Coefficients}
summary(aft_accidents_weibull)
summary(aft_cerebro_weibull)
summary(aft_covid_weibull)
summary(aft_heart_weibull)
summary(aft_neoplasms_weibull)

exp(coef(aft_accidents_weibull))
exp(coef(aft_cerebro_weibull))
exp(coef(aft_covid_weibull))
exp(coef(aft_heart_weibull))
exp(coef(aft_neoplasms_weibull))
```
```{r Compare with actual Values, example Heart Disease}
#To analyse differences between predictions
predict_diff <- function(sex1, sex2, urb) {
  predict(aft_heart_weibull, newdata = data.frame(Sex = sex1,  Urbanization = urb)) -
    predict(aft_heart_weibull, newdata = data.frame(Sex = sex2, Urbanization = urb))
}
# Example:
predict_diff("Male", "Female", "Small Metro")

#Get expected Age:
predicted_age <- function(sex, urb, data = aft_data) {
predict(aft_heart_weibull, newdata = data.frame(
  Sex = "Female",
  Urbanization = "Large Central Metro"
))
}
#Validate on actual data:
get_observed_age_summary <- function(sex, urb, data = aft_data) {
  data %>%
    filter(
      Sex == sex,
      Urbanization == urb
    ) %>%
    summarise(
      mean_age = mean(Age),
      median_age = median(Age),
      count = n()
    )
}
predicted_age(sex, urb, data = aft_data)
get_observed_age_summary(sex = "Female", urb = "Small Metro")

#not that great, test with other distributions and adjusted Population
```

```{r Residuals}
# Define models using new variable names
models <- list(
  "Accidents (unintentional injuries)" = aft_accidents_weibull,
  "Cerebrovascular diseases" = aft_cerebro_weibull,
  "COVID-19" = aft_covid_weibull,
  "Diseases of heart" = aft_heart_weibull,
  "Malignant neoplasms" = aft_neoplasms_weibull
)

# Loop through each model and generate diagnostic plots
for (cod in names(models)) {
  model <- models[[cod]]
  data_subset <- aft_data %>% filter(CoD == cod)
  
  # Residuals
  raw_log_resid <- log(data_subset$Age) - model$linear.predictors
  std_log_resid <- raw_log_resid / model$scale
  dev_resid <- residuals(model)
  pred_age <- predict(model, type = "response")
  obs_age <- data_subset$Age

  # Plot layout
  par(mfrow = c(2, 3))
  
  # 1. Deviance residuals
  plot(dev_resid, main = paste("Deviance Residuals -", cod), ylab = "Residual")
  abline(h = 0, col = "red", lty = 2)
  
  # 2. Standardized log-residuals
  plot(std_log_resid, main = paste("Standardized Log-Residuals -", cod), ylab = "Residual")
  abline(h = 0, col = "red", lty = 2)
  
  # 3. QQ Plot of Log-Residuals
  qqnorm(raw_log_resid, main = paste("QQ Plot of Log-Residuals -", cod))
  qqline(raw_log_resid)
  
  # 4. QQ Plot of Deviance Residuals
  qqnorm(dev_resid, main = paste("QQ Plot of Deviance Residuals -", cod))
  qqline(dev_resid)
  
  # 5. Predicted vs Observed
  plot(pred_age, obs_age,
       xlab = "Predicted Age", ylab = "Observed Age",
       main = paste("Predicted vs Observed -", cod))
  abline(0, 1, col = "red")
  
  # 6. Histogram of Log-Residuals
  hist(raw_log_resid, breaks = 50, main = paste("Histogram -", cod), xlab = "Log-Residual")
  
  # Pause between plots
  readline(prompt = paste0("Press [Enter] to continue to ", cod, "..."))
}
#couldnt fix it, still bad. Even with other distributions
```








