---
title: "Multinomial"
---

#labRiaies
```{r}

# install.packages("readr")     
# install.packages("dplyr")     
# install.packages("stringr")   
# install.packages("VGAM")   
library(readr)
library(dplyr)
library(stringr)
library(nnet)
library(knitr)
library(VGAM)
library(broom)
library(dplyr)
library(tidyr)
library(knitr)
library(here)
```




#Read data 
```{r}
# Reading each CSV file 
data_mult <- read_csv(here("All_Causes_Age_Groups10.csv"))
```

# Data Preparation and Overview
```{r}

# Filter rows where Cause is NOT "All Deaths"
data_mult_filtered <- data_mult %>%
  filter(Cause != "All Deaths")

# Check structure of dataset
glimpse(data_mult)


# Identify categorical (character) columns
categorical_columns <- sapply(data_mult, is.character)

# Extract unique values from each categorical column
unique_values <- lapply(data_mult[, categorical_columns], unique)

# Print unique values for each categorical variable
for (col in names(unique_values)) {
  cat("Unique values in variable:", col, "\n")
  print(unique_values[[col]])
  cat("\n")
}
```



# Data Preparation
```{r}
# Filter the data: keep only rows where key variables are not missing, deaths are valid, and age group is in the desired list

data_mult_filtered <- data_mult_filtered %>%
  filter(
    !is.na(Cause),
    !is.na(Sex),
    !is.na(Education),
    !is.na(`Ten-Year Age Groups`),
    !is.na(Deaths),
    is.finite(Deaths),
    Deaths > 0,
    `Ten-Year Age Groups` %in% c("25-34 years", "35-44 years", "45-54 years",
                                 "55-64 years", "65-74 years", "75-84 years", "85+ years")
  ) %>%
  mutate(
    Deaths = round(as.numeric(Deaths)), # Convert Deaths to numeric and round the values
    Cause = relevel(factor(Cause), ref = "Others") # Convert Cause to factor and set "Others" as the reference level
  )

# Remove the 'adjusted_population' column from the dataset
data_mult_filtered <- data_mult_filtered %>%
  select(-adjusted_population)

# Display summary statistics of the filtered dataset
summary(data_mult_filtered)


# Check if there are any missing values remaining in the dataset
any(is.na(data_mult_filtered))

```







# Model
Sex + Education + `Ten-Year Age Groups`

#Шансы смерти от причины внутри всех смертей \\ Это модель структуры смертности, а не риска смерти.
#Chance of death from cause within all deaths \\ This is a mortality structure model, not a risk of death.

```{r}
# Check levels
levels(data_mult_filtered$Cause)
```


#Model : Education

```{r}

model_education <- vglm(Cause ~ Education,
              family = multinomial(refLevel = "Others"),
              weights = Deaths,
              data = data_mult_filtered)

summary(model_education)
```





#Model : Sex + Education + Ten-Year Age Groups

```{r}
model_sex_education_age <- vglm(Cause ~ Sex + Education + `Ten-Year Age Groups`,
              family = multinomial(refLevel = "Others"),
              weights = Deaths,
              data = data_mult_filtered)

summary(model_sex_education_age)
```




#Model : Sex + Education + Ten-Year Age Groups + Interaction
```{r}
model_interaction <- vglm(Cause ~ Sex + Education + `Ten-Year Age Groups`+ Education*Sex + Education*`Ten-Year Age Groups`,
              family = multinomial(refLevel = "Others"),
              weights = Deaths,
              data = data_mult_filtered)

summary(model_interaction)
```




# Table of cofficients of Model with Intraction 
```{r}
library(dplyr)
library(tidyr)
library(stringr)

# 1. Extract coefficients
coef_table <- coef(summary(model_interaction))
coef_df <- as.data.frame(coef_table)
coef_df$Combined <- rownames(coef_df)

# 2. Remove Intercept
coef_df <- coef_df %>% filter(!str_detect(Combined, "\\(Intercept\\)"))

# 3. Extract CauseLevel (the number after the last ":") and Predictor
coef_df <- coef_df %>%
  mutate(
    CauseID = str_extract(Combined, "[^:]+$"),
    Predictor = str_replace(Combined, ":[^:]+$", "")
  )

# 4. Add confidence intervals
colnames(coef_df)[1:2] <- c("Estimate", "StdError")
coef_df <- coef_df %>%
  mutate(
    Lower = Estimate - 1.96 * StdError,
    Upper = Estimate + 1.96 * StdError
  )

# 5. Display top 5 coefficients for each CauseID
top5_df <- coef_df %>%
  group_by(CauseID) %>%
  arrange(desc(Estimate)) %>%
  slice_head(n = 5) %>%
  ungroup()

# 6. Print the table
print(top5_df)

```



# Gafik of coefieccients of Model with Intractions

```{r}
library(ggplot2)

# 1. Create a mapping vector
cause_labels <- c(
  "1" = "Accidents",
  "2" = "Cerebrovascular Diseases",
  "3" = "COVID",
  "4" = "Diseases of Heart",
  "5" = "Neoplasms"
)

# 2. Convert CauseID to a factor with labels
top5_df <- top5_df %>%
  mutate(Cause = recode(CauseID, !!!cause_labels))

# 1. Plot top-5 predictors (including interactions) for each cause
unique_causes <- unique(top5_df$CauseID)

# Use Cause in the plot title
for (cid in unique(top5_df$Cause)) {
  plot_data <- filter(top5_df, Cause == cid)
  
  p <- ggplot(plot_data, aes(x = reorder(Predictor, Estimate), y = Estimate)) +
    geom_pointrange(aes(ymin = Lower, ymax = Upper), color = "black", size = 0.8) +
    coord_flip() +
    theme_bw(base_size = 20) +  # Increase base text size
    theme(
      plot.title = element_text(size = 24, face = "bold"),
      axis.title.x = element_text(size = 20),
      axis.title.y = element_text(size = 20),
      axis.text = element_text(size = 18)
    ) +
    labs(
      title = paste("Top 5 Predictors (including interactions) for Cause:", cid),
      x = "Predictor",
      y = "Coefficient (log-odds)"
    )
  
  print(p)
}
```






# Model Compaison AIC\BIC
```{r}
# Собираем метрики
model_comparison <- data.frame(
  Model = c("1: Education",
            "2: Education + Age + Sex",
             "3: with Interaction"),
  # LogLikelihood = c(logLik(model_education),
  #                   logLik(model_sex_education_age),
  #                   logLik(model_interaction)),
  AIC = c(AIC(model_education),
          AIC(model_sex_education_age),
          AIC(model_interaction)),
  BIC = c(BIC(model_education),
          BIC(model_sex_education_age),
          BIC(model_interaction))
)

# Округляем для наглядности
#model_comparison$LogLikelihood <- round(model_comparison$LogLikelihood, 2)
model_comparison$AIC <- round(model_comparison$AIC, 2)
model_comparison$BIC <- round(model_comparison$BIC, 2)

# Вывод таблицы
model_comparison

```


# END --------------------------------------------------------------------------------------------


















#Other 


```{r}
# model_interaction_v2 <- vglm(Cause ~  Education + Education*Sex + Education*`Ten-Year Age Groups`,
#               family = multinomial(refLevel = "Others"),
#               weights = Deaths,
#               data = data_mult_filtered)
# 
# summary(model_interaction_v2)
```





#Bayes model
```{r}
# # Убедись, что установлены нужные пакеты
# #install.packages("brms")        # Если не установлен
# library(brms)
# 
# 
# 
# # Байесовская мультиномиальная логистическая регрессия
# bayes_model <- brm(
#   formula = Cause ~ Education,
#   data = data_mult_filtered,
#   family = categorical(link = "logit"),
# #  weights = Deaths,
#   chains = 4,
#   cores = 4,
#   iter = 2000,
#   seed = 123
# )
# 
# # Сводка модели
# summary(bayes_model)
# 
# # Визуализация коэффициентов
# plot(bayes_model)

```





#fis woking model
```{r}
# model <- vglm(Cause ~ Sex + Education ,
#               family = multinomial(refLevel = "Others"),
#               weights = Deaths,
#               data = data_mult_filtered)
# 
# summary(model)
```


#penalised v1 
```{r}
# model <- vglm(Cause ~  Sex + Education, family = multinomial(refLevel = "Others"), data = data_mult_filtered,
#               weights = Deaths, maxit = 50, control = vglm.control(trace = TRUE), penalty = TRUE)
# 
# 
# 
# summary(model)

```

#penalised v2 
```{r}
# library(glmnet)

# # Подготовка данных
# X <- model.matrix(Cause ~ Sex + Education, data = data_mult_filtered)[, -1]
# y <- factor(data_mult_filtered$Cause)
# 
# # Веса: Deaths
# w <- data_mult_filtered$Deaths
# 
# # Обучение с регуляризацией (L2 по умолчанию)
# cv_fit <- cv.glmnet(X, y, family = "multinomial", weights = w)
# 
# # Коэффициенты при оптимальном lambda
# coef(cv_fit, s = "lambda.min")
```



# penalised v3
```{r}
# install.packages("pemultinom")
# 
# library(pemultinom)
# 
# # Подготовка данных: преобразование предикторов в матрицу и отклика в фактор
# x <- model.matrix(~  Education, data = data_mult_filtered)[, -1]
# y <- as.factor(data_mult_filtered$Cause)
# 
# # Обучение модели с кросс-валидацией для выбора параметра регуляризации
# fit <- cv.pemultinom(x = x, y = y, nfolds = 5)
# 
# # Извлечение коэффициентов модели при оптимальном значении lambda
# beta <- fit$beta.min
# 
# # Получение предсказаний на новых данных (пример)
# predictions <- predict_pemultinom(fit, newx = x)

```