```{r Packages, echo = FALSE}
library(survival)
library(survminer)
library(dplyr)
library(tidyr)
library(readr)
library(here)
library(nnet)
library(ggplot2)
library(MASS)
library(flexsurv)
library(timereg)
library(mgcv)
library(cmprsk)
library(broom)
library(splines)
library(purrr)
library(pscl)
options(scipen = 999)
```

```{r Loading Data & Adjusting}
file_path <- here("Datasets", "Population Data Merge", "All_Causes_Combined.csv")
df <- read.csv(file_path)
df <- df[, c("Cause", "AGE", "Sex", "Education", "Deaths", "population")]
df <- df %>%
  mutate(Cause = as.factor(Cause)) %>%
  mutate(Sex = as.factor(Sex)) %>%
  mutate(Education = as.factor(Education)) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  mutate(population = round(population))
df %>% filter(is.na(population))
df$Education <- factor(df$Education, levels = c(
  "8th grade or less",
  "9th through 12th grade with no diploma",
  "High school graduate or GED completed",
  "Some college credit, but not a degree",
  "Associate degree (AA,AS)",
  "Bachelor’s degree (BA, AB, BS)",
  "Master’s degree (MA, MS, MEng, MEd, MSW, MBA)",
  "Doctorate (PhD, EdD) or Professional Degree (MD, DDS, DVM, LLB, JD)"
), ordered = FALSE)
df$Cause <- relevel(df$Cause, ref = "Others")
```
#Problem: There are Deaths for Groups without a Population. This happens primarily in groups which we can expect to be quite rare. (Young People with above average Education level)
# However, due to our research Goal: analysing Mortality by Educational Goal we can remove People under 25 anyways, becasue they would just bias our results due to the fact, that we cannot assume future highest educational attainment for people presumably in the midst of their education
Keep in Mind, we are also not analysing People over 79 due to coding issues from the IPUMS Dataset
```{r}

```

```{r}
scale_factor <- 100000
reduced_deaths <- df %>%
    mutate(Deaths = (Deaths / population)*scale_factor) %>%
    mutate(Deaths = round(Deaths)) %>%
    filter(!is.na(Deaths))

expanded_deaths <- reduced_deaths %>%
  uncount(weights = Deaths) %>%
  mutate(status = 1)

aft_data <- expanded_deaths %>%
  filter(!is.na(AGE) & AGE > 0)
```

```{r Models, all interactions}
# Weibull
aft_weibull <- survreg(Surv(AGE) ~ Sex * Cause + Sex * Education + Cause*Education, data = aft_data, dist = "weibull")

# Log-normal
aft_lognormal <- survreg(Surv(AGE) ~ Sex * Cause + Sex * Education + Cause*Education, data = aft_data, dist = "lognormal")

# Log-logistic
aft_loglogistic <- survreg(Surv(AGE) ~ Sex * Cause + Sex * Education + Cause*Education, data = aft_data, dist = "loglogistic")

# Exponential
aft_exponential <- survreg(Surv(AGE) ~ Sex * Cause + Sex * Education + Cause*Education, data = aft_data, dist = "exponential")

#show the reference Category
levels(aft_data$Sex)[1]
levels(aft_data$Cause)[1]
levels(aft_data$Education)[1]

#lognormal has the correct intercept
```

```{r Best Model}
#compare with AIC
AIC(aft_weibull, aft_lognormal, aft_loglogistic, aft_exponential)

#compare log.likelihoods
logLik(aft_weibull)
logLik(aft_lognormal)
logLik(aft_loglogistic)
logLik(aft_exponential)

#Weibull is best
```

```{r}
aft_weibull2 <- survreg(Surv(AGE) ~ Sex * Education + Cause * Education, data = aft_data, dist = "weibull")
```

```{r}
summary(aft_weibull2)
exp(coef(summary(aft_weibull2)))
```

```{r}
# --- Extract model predictions and residuals ---
model_data <- aft_weibull2$model

model_results <- aft_data %>%
  mutate(
    predicted = predict(aft_weibull2, type = "response"),
    resid = residuals(aft_weibull2, type = "response"),
    std_resid = resid / sqrt(predicted),
    outlier = abs(std_resid) > 4
  ) %>%
  arrange(predicted) %>%
  mutate(index = row_number()) %>%
  group_by(Cause) %>%
  mutate(index_within_CoD = row_number()) %>%
  ungroup()


# --- Merge with original data ---
aft_data$row_id <- as.numeric(rownames(aft_data))

df_with_results <- aft_data %>%
  mutate(
    predicted = predict(aft_weibull2, type = "response"),
    resid = residuals(aft_weibull2, type = "response"),
    std_resid = resid / sqrt(predicted),
    outlier = abs(std_resid) > 4
  ) %>%
  arrange(predicted) %>%
  mutate(index = row_number()) %>%
  group_by(Cause) %>%
  mutate(index_within_CoD = row_number()) %>%
  ungroup()

# --- Add standardized residuals, flags, and sorted indices ---
df_with_results <- df_with_results %>%
  mutate(
    std_resid = resid / sqrt(predicted),
    outlier = abs(resid) > 4
  ) %>%
  arrange(predicted) %>%
  mutate(index = row_number()) %>%
  group_by(Cause) %>%
  mutate(index_within_CoD = row_number()) %>%
  ungroup()


# --- QQ Plot of Pearson Residuals ---
qqnorm(df_with_results$resid, main = "QQ Plot of Pearson Residuals")
qqline(df_with_results$resid)

# --- Standardized Residuals vs Sorted Index ---
plot(df_with_results$index, df_with_results$std_resid,
     main = "Standardized Pearson Residuals",
     ylab = "Standardized Residuals", xlab = "Sorted Observation Index", pch = 20, cex = 0.5)
abline(h = 0, col = "red", lty = 2)
abline(h = c(-2, 2), col = "gray", lty = 3)

# --- Residuals by Sex ---
ggplot(df_with_results, aes(x = index, y = resid, color = Sex)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Pearson Residuals by Sex", x = "Sorted Observation Index", y = "Residual")

# --- Residuals by Cause (shared axis) ---
ggplot(df_with_results, aes(x = index, y = resid, color = Cause)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Pearson Residuals by Cause of Death (Shared X)", x = "Sorted Observation Index", y = "Residual")

# --- Residuals by Cause (faceted) ---
ggplot(df_with_results, aes(x = index_within_CoD, y = resid)) +
  geom_point(alpha = 0.4, size = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ Cause, scales = "free_x") +
  theme_minimal() +
  labs(title = "Pearson Residuals Faceted by Cause of Death", x = "Index (within Cause)", y = "Residual")

# --- Histogram of Residuals ---
hist(df_with_results$resid, breaks = 50,
     main = "Histogram of Pearson Residuals", xlab = "Residual")

```



