```{r Packages}
library(survival)
library(survminer)
library(dplyr)
library(readr)
library(here)
library(nnet)
library(ggplot2)
library(MASS)
library(flexsurv)
library(timereg)
library(mgcv)
library(cmprsk)
library(broom)
library(splines)
library(purrr)
library(pscl)
options(scipen = 999)
```

```{r Loading Data & Adjusting}
file_path <- here("Datasets", "Population Data Merge", "All_Causes_Combined.csv")
df <- read.csv(file_path)
df <- df[, c("Cause", "AGE", "Sex", "Education", "Deaths", "population")]
df <- df %>%
  mutate(Cause = as.factor(Cause)) %>%
  mutate(Sex = as.factor(Sex)) %>%
  mutate(Education = as.factor(Education)) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  mutate(population = round(population))
df$Education <- factor(df$Education, levels = c(
  "8th grade or less",
  "9th through 12th grade with no diploma",
  "High school graduate or GED completed",
  "Some college credit, but not a degree",
  "Associate degree (AA,AS)",
  "Bachelor’s degree (BA, AB, BS)",
  "Master’s degree (MA, MS, MEng, MEd, MSW, MBA)",
  "Doctorate (PhD, EdD) or Professional Degree (MD, DDS, DVM, LLB, JD)"
), ordered = FALSE)
df$Cause <- relevel(df$Cause, ref = "Others")
```
#Problem: There are Deaths for Groups without a Population. This happens primarily in groups which we can expect to be quite rare. (Young People with above average Education level)
# However, due to our research Goal: analysing Mortality by Educational Goal we can remove People under 25 anyways, becasue they would just bias our results due to the fact, that we cannot assume future highest educational attainment for people presumably in the midst of their education
Keep in Mind, we are also not analysing People over 79 due to coding issues from the IPUMS Dataset
```{r}
df <- df %>% filter(AGE > 24)
```

```{r Loading Data & Adjusting, suppressed Values included}
file_path <- here("Datasets", "Population Data Merge", "All_Causes_Combined_sup_incl.csv")
df2 <- read.csv(file_path)
df2 <- df2[, c("Cause", "AGE", "Sex", "Education", "Deaths", "population")]
df2 <- df2 %>%
  mutate(Cause = as.factor(Cause)) %>%
  mutate(Sex = as.factor(Sex)) %>%
  mutate(Education = as.factor(Education)) %>%
  mutate(Deaths = as.numeric(Deaths)) %>%
  mutate(population = round(population))
df2$Education <- factor(df2$Education, levels = c(
  "8th grade or less",
  "9th through 12th grade with no diploma",
  "High school graduate or GED completed",
  "Some college credit, but not a degree",
  "Associate degree (AA,AS)",
  "Bachelor’s degree (BA, AB, BS)",
  "Master’s degree (MA, MS, MEng, MEd, MSW, MBA)",
  "Doctorate (PhD, EdD) or Professional Degree (MD, DDS, DVM, LLB, JD)"
), ordered = FALSE)
df2$Cause <- relevel(df2$Cause, ref = "Others")
df2 <- df2 %>% filter(AGE > 24)
```
#Problem: There are Deaths for Groups without a Population. This happens primarily in groups which we can expect to be quite rare. (Young People with above average Education level)
# However, due to our research Goal: analysing Mortality by Educational Goal we can remove People under 25 anyways, becasue they would just bias our results due to the fact, that we cannot assume future highest educational attainment for people presumably in the midst of their education
Keep in Mind, we are also not analysing People over 79 due to coding issues from the IPUMS Dataset
```{r}
df <- df %>% filter(AGE > 24)
```

```{r Deaths per Age & Education level}
causes <- unique(df$Cause)
for (cause in causes) {
  p <- ggplot(subset(df, Cause == cause), aes(x = AGE, y = Deaths, color = Education)) +
    geom_point(alpha = 0.7) +
    labs(title = paste("Deaths by Age and Sex –", cause),
         x = "Age",
         y = "Number of Deaths") +
    theme_minimal()
  
  print(p)
}
```

```{r Deaths per Education per Capita}
for (cause in unique(df$Cause)) {
  df_cause <- subset(df, Cause == cause & population > 0)
  
  p <- ggplot(df_cause, aes(x = Education, y = Deaths / population, fill = Sex)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = paste("Normalized Death Rate by Education and Sex –", cause),
      x = "Education Level",
      y = "Deaths per Capita"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))  # improves x-axis readability
  
  print(p)
}
```

```{r Deaths per Capita over Age & Education level}

for (cause in causes) {
  df_cause <- subset(df, Cause == cause & population > 0)
  
  p <- ggplot(df_cause, aes(x = AGE, y = Deaths / population, color = Education)) +
    geom_point(alpha = 0.7) +
    labs(
      title = paste("Normalized Death Rate by Age and Education –", cause),
      x = "Age",
      y = "Deaths per Capita"
    ) +
    theme_minimal()
  
  print(p)
}
```

```{r Deaths per Capita over Age & Education level, seperated by Sex}
for (cause in unique(df$Cause)) {
  df_cause <- subset(df, Cause == cause & population > 0)
  
  p <- ggplot(df_cause, aes(x = AGE, y = Deaths / population, color = Education)) +
    geom_point(alpha = 0.7) +
    facet_wrap(~ Sex) +
    labs(
      title = paste("Normalized Death Rate by Age, Education, and Sex –", cause),
      x = "Age",
      y = "Deaths per Capita"
    ) +
    theme_minimal()
  
  print(p)
}
```
From the plots we can read that Higher Educational Attainment correlates with lower Death rates. 

```{r Poisson regression & Dispersion Calculation}
poisson_model <- glm(Deaths ~ AGE + Sex + Education + Cause + offset(log(population)),
                     data = df,
                     family = poisson(link = "log"))

summary(poisson_model)
# Calculate dispersion
dispersion <- sum(residuals(poisson_model, type = "pearson")^2) / poisson_model$df.residual
dispersion #should be 1 --> significant overdispersion
```

```{r Negative Binomial Model}
nb_model <- glm.nb(Deaths ~ AGE + Sex + Education + Cause + offset(log(population)),
                    data = df,
                  )
summary(nb_model)
exp(coefficients(nb_model))
#Should we use a Zero-inflated Model?
mean(df$Deaths == 0)                        # Observed zero rate
mean(predict(nb_model, type = "response") < 0.5)  # Predicted near-zero rate
nrow(df %>% filter(Deaths==0))
#Even though Model will not predict Deaths that are 0, we only have 0.6% Zero-Values so using a Zero-Inflated Negative Binomial Model is not justified
```

```{r Improving the Model}
nb_final <- glm.nb(Deaths ~ ns(AGE, 3) * Sex + Education * Sex + Cause * Sex + Education * Cause + ns(AGE, 3) * Cause + offset(log(population)), 
               data = df)
summary(nb_final)
exp(coefficients(nb_final))
```

```{r Impriving Model sup incl}
nb_final <- glm.nb(Deaths ~ ns(AGE, 3) * Sex + Education * Sex + Cause * Sex + Education * Cause + ns(AGE, 3) * Cause + offset(log(population)), 
               data = df2)
summary(nb_final)
exp(coefficients(nb_final)) ###small changes, barely noticeable. When rounding we might be off for 1 Percentage Point
```


```{r Testing the Model}
# Remove Age splines
m_no_spline <- update(nb_final, . ~ . - ns(AGE, 3) - ns(AGE, 3):Sex - ns(AGE, 3):Cause)
anova(m_no_spline, nb_final, test = "Chisq")

# Remove Age:Sex interaction
m_no_age_sex <- update(nb_final, . ~ . - ns(AGE, 3):Sex)
anova(m_no_age_sex, nb_final, test = "Chisq")

# Remove Education:Sex
m_no_edu_sex <- update(nb_final, . ~ . - Education:Sex)
anova(m_no_edu_sex, nb_final, test = "Chisq")


AIC(nb_final, m_no_spline, m_no_age_sex, m_no_edu_sex)
```


```{r Models split by Cause for easier Interpretability}
cause_list <- unique(df$Cause)

# Create an empty list to store models
cause_models <- list()

# Loop over each cause and fit the model
for (cause in cause_list) {
  
  # Subset data for this cause
  df_subset <- df[df$Cause == cause, ]
  
  # Fit the model (excluding Cause terms since we’re conditioning on one cause)
  model <- glm.nb(
    Deaths ~ ns(AGE, 3) * Sex + Education * Sex + offset(log(population)),
    data = df_subset
  )
  
  # Store the model in the list
  cause_models[[cause]] <- model
}
cause_models[["Diseases of Heart"]]
summary(cause_models[["Diseases of Heart"]])
```


```{r}
# --- Extract model predictions and residuals ---
model_data <- nb_final$model
model_results <- data.frame(
  row_id = as.numeric(rownames(model_data)),
  predicted = predict(nb_final, type = "response"),
  resid = residuals(nb_final, type = "pearson")
)

# --- Merge with original data ---
df$row_id <- as.numeric(rownames(df))
df_with_results <- left_join(df, model_results, by = "row_id")

# --- Add standardized residuals, flags, and sorted indices ---
df_with_results <- df_with_results %>%
  mutate(
    std_resid = resid / sqrt(predicted),
    outlier = abs(resid) > 4
  ) %>%
  arrange(predicted) %>%
  mutate(index = row_number()) %>%
  group_by(Cause) %>%
  mutate(index_within_CoD = row_number()) %>%
  ungroup()

# --- Predicted vs Observed (log-log) ---
log_plot_data <- df_with_results %>% filter(predicted > 0, Deaths > 0)
plot(log_plot_data$predicted, log_plot_data$Deaths,
     log = "xy",
     xlab = "Predicted Deaths (log scale)", ylab = "Observed Deaths (log scale)",
     main = "Predicted vs. Observed Deaths")
abline(0, 1, col = "red")

# --- QQ Plot of Pearson Residuals ---
qqnorm(df_with_results$resid, main = "QQ Plot of Pearson Residuals")
qqline(df_with_results$resid)

# --- Standardized Residuals vs Sorted Index ---
plot(df_with_results$index, df_with_results$std_resid,
     main = "Standardized Pearson Residuals",
     ylab = "Standardized Residuals", xlab = "Sorted Observation Index", pch = 20, cex = 0.5)
abline(h = 0, col = "red", lty = 2)
abline(h = c(-2, 2), col = "gray", lty = 3)

# --- Residuals by Sex ---
ggplot(df_with_results, aes(x = index, y = resid, color = Sex)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Pearson Residuals by Sex", x = "Sorted Observation Index", y = "Residual")

# --- Residuals by CoD (mixed on one axis) ---
ggplot(df_with_results, aes(x = index, y = resid, color = Cause)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Pearson Residuals by Cause of Death (Shared X)", x = "Sorted Observation Index", y = "Residual")

# --- Residuals by CoD (faceted per group) ---
ggplot(df_with_results, aes(x = index_within_CoD, y = resid)) +
  geom_point(alpha = 0.4, size = 0.6, color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~ Cause, scales = "free_x") +
  theme_minimal() +
  labs(title = "Pearson Residuals Faceted by Cause of Death", x = "Index (within CoD)", y = "Residual")

# --- Histogram of Residuals ---
hist(df_with_results$resid, breaks = 50,
     main = "Histogram of Pearson Residuals", xlab = "Residual")
```


!!!! this does not tell us, how does education affect whether you die? But tells us: How does Education affect what you die of?  Whats the most likely cause of death?
```{r Multinomial Model}

model <- nnet::multinom(Cause ~ Education + Sex + AGE, data = df, weights = Deaths)

newdata <- data.frame(
  Sex = "Male",
  AGE = 65,
  Education = unique(df$Education)
)

predicted_probs <- predict(multi_model, newdata = newdata, type = "probs")
prediction_output <- cbind(newdata, as.data.frame(predicted_probs))

# Step 2: Observed proportions for 65-year-old males
observed_65M <- df %>%
  filter(Sex == "Male", AGE == 65) %>%
  group_by(Cause) %>%
  summarise(Deaths = sum(Deaths), .groups = "drop") %>%
  mutate(Observed_Prob = Deaths / sum(Deaths))

# Step 3: Comparison for a single education level (e.g., "High school graduate or GED completed")
edu_focus <- "High school graduate or GED completed"

# Extract predicted row
pred_row <- prediction_output %>%
  filter(Education == edu_focus) %>%
  dplyr::select(-Sex, -AGE, -Education) %>%
  pivot_longer(cols = everything(), names_to = "Cause", values_to = "Predicted_Prob")

# Merge predicted and observed
comparison <- observed_65M %>%
  left_join(pred_row, by = "Cause")

# View result
print(comparison)

summary(model)

model_summary <- tidy(model)
exp(model_summary)
```






